---
title: "ANN Benchmarks"
author: "Chip Lynch"
date: "2/23/2021"
output: pdf_document
---

## ANN Benchmarks

```{html echo=FALSE}
<!--
## Kentucky 2d Geodesic 150000 points:
(venv38) chipmonkey@chipbuntu340 ~/repos/ann-benchmarks (feature/trilateration_index)$ ./run_trilat.sh 
2021-03-16 18:55:43,367 - annb - INFO - running only ball
2021-03-16 18:55:43,690 - annb - INFO - Order: [Definition(algorithm='ball', constructor='BallTree', module='ann_benchmarks.algorithms.balltree', docker_tag='ann-benchmarks-sklearn', arguments=['geodesic', 400], query_argument_groups=[], disabled=True), Definition(algorithm='ball', constructor='BallTree', module='ann_benchmarks.algorithms.balltree', docker_tag='ann-benchmarks-sklearn', arguments=['geodesic', 20], query_argument_groups=[], disabled=True), Definition(algorithm='ball', constructor='BallTree', module='ann_benchmarks.algorithms.balltree', docker_tag='ann-benchmarks-sklearn', arguments=['geodesic', 10], query_argument_groups=[], disabled=True), Definition(algorithm='ball', constructor='BallTree', module='ann_benchmarks.algorithms.balltree', docker_tag='ann-benchmarks-sklearn', arguments=['geodesic', 1000], query_argument_groups=[], disabled=True), Definition(algorithm='ball', constructor='BallTree', module='ann_benchmarks.algorithms.balltree', docker_tag='ann-benchmarks-sklearn', arguments=['geodesic', 200], query_argument_groups=[], disabled=True), Definition(algorithm='ball', constructor='BallTree', module='ann_benchmarks.algorithms.balltree', docker_tag='ann-benchmarks-sklearn', arguments=['geodesic', 100], query_argument_groups=[], disabled=True), Definition(algorithm='ball', constructor='BallTree', module='ann_benchmarks.algorithms.balltree', docker_tag='ann-benchmarks-sklearn', arguments=['geodesic', 40], query_argument_groups=[], disabled=True)]
Trying to instantiate ann_benchmarks.algorithms.balltree.BallTree(['geodesic', 400])
got a train set of size (135000 * 2)
got 15000 queries
Built index in 242.90340185165405
Index size:  2852.0
Running query argument group 1 of 1...
Run 1/5...
Processed 1000/15000 queries...
Processed 2000/15000 queries...
Processed 3000/15000 queries...
Processed 4000/15000 queries...
Processed 5000/15000 queries...
Processed 6000/15000 queries...
Processed 7000/15000 queries...
Processed 8000/15000 queries...
Processed 9000/15000 queries...
Processed 10000/15000 queries...
Processed 11000/15000 queries...
Processed 12000/15000 queries...
Processed 13000/15000 queries...
Processed 14000/15000 queries...
Stoped at Wed Mar 17 16:45:17 PDT 2021  (~22 hours into the run)
So like 5 seconds per query.  (?!)
-->
```


As hoped, our Trilateration algorithm really shines when applying the complex but accurate geodesic distance function.  The Trilateration algorithm is over 30 times faster than the next best candidate (the Ball Tree algorithm with leaf_size=10).  The Brute Force algorithm is unbearably slow here, which is expected since it should be calling the expensive distance function $n^2$ or $150,000^2=22,500,000,000$ (22.5 billion) times... more than any other algorithm by far.  Recall that our test of 5000 calls to a single geodetic function took $\approx1.2$ seconds, so we'd expect 22.5 billion to take 5.4 million seconds, or about 62 days on similar hardware.  The other algorithms ran to completion.

```{html echo=FALSE}
<!--
                         Name           Accuracy          QPS   Index Time
  0:         BallTree(leaf_size=10)        1.000        2.249      258.140
  1:        BallTree(leaf_size=100)        1.000        0.683      207.249
  2:       BallTree(leaf_size=1000)        1.000        0.182      153.778
  3:           TrilaterationExpand2        1.000        1.335       59.262
  4:                   BruteForce()        1.000*       0.050*       0.001*
  5:           KDTree(leaf_size=10)        1.000        0.099        0.125
  6:          KDTree(leaf_size=100)        1.000        0.096        0.106
  7:         KDTree(leaf_size=1000)        1.000        0.089        0.085
  8:                  Trilateration        1.000       73.119       58.857
  9:                   TrilatExpand        1.000        0.364       58.737
* BruteForce algorithm did not run to completion due to exceeding time limits
-->
```

```{r ann_geodesic_25, echo=FALSE, fig.cap="Queries Per Second for Geodesic (Q=150,000; higher is better)", fig.width=6, fig.height=3}
library(knitr)
times <- data.frame(Algorithm=c("BallTree(leaf_size=10)", "BallTree(leaf_size=100)", "BallTree(leaf_size=1000)",
                            "TrilaterationExpand2", "BruteForce", "KDTree(leaf_size=10)", "KDTree(leaf_size=100)",
                            "KDTree(leaf_size=1000)", "Trilateration", "TrilaterationExpand"),
                    QPS=c(2.249, 0.683, 0.182, 1.335, 0.050, 0.099, 0.096, 0.089, 73.119, 0.364),
                    IndexBuild=c(258.140, 207.249, 153.778, 59.262, 0.001, 0.125, 0.106, 0.085, 58.875, 58.737))
times <- times[order(times$Algorithm),]
# kable(times[,1:3], caption="Queries Per Second and Time to Build Indexes for Geodesic (Q=150,000)")

p <- ggplot(data=times, aes(x=Algorithm, y=QPS)) +
  geom_bar(stat="identity", col='black', fill='grey') +
  coord_flip()
p
```


### Random 20-dimension euclidean distance:

The ann-benchmark tool includes support for a randomly generated 20-dimension euclidean dataset, which is one of its most basic tests.  We note that our performance here is abysmal for the initial "Trilateration" algorithm, even failing to beat the brute force approach.  This seems to be due to the overhead we incur determining which candidates to test next.  Remember that we traded a number of subtractions and some array navigation in exchange for fewer distance function calls.  In this case, when the distance function itself is extremely fast, that overhead is a net loss.

It is worth noting that we tuned the two expansion based algorithms here as well.  "TrialterationExpand" performs far better than the stock Trilateration algorithm, but still just below the Brute Force algorithm.  The "TrilaterationExpand2" algorithm, however, is actually competitive here, more than doubling the queries per second of the Brute Force approach, and reaching more than 60% as fast as some tree algorithms.

For Euclidean distances, however, we cannot recommend our algorithms against competitors.

```{html echo=FALSE}
<!--
writing output to results/random-xs-20-euclidean.png
  0:                                          TrilaterationExpand        1.000      401.899
  1:                                                Trilateration        1.000       25.971
  2:                                          TrilaterationApprox        0.585       71.236
  3:                                         TrilaterationExpand2        1.000     1411.711
  4:                                       BallTree(leaf_size=40)        1.000     3040.686
  5:                                       BallTree(leaf_size=10)        1.000     3974.894
  6:                                     BallTree(leaf_size=1000)        1.000     2224.275
  7:                                      BallTree(leaf_size=100)        1.000     2561.925
  8:                                      BallTree(leaf_size=400)        1.000     2265.526
  9:                                      BallTree(leaf_size=200)        1.000     2352.948
 10:                                       BallTree(leaf_size=20)        1.000     3532.328
 11:                                                 BruteForce()        1.000      670.163
 -->
```

```{r random-20-euclidean, echo=FALSE, results ='asis', fig.cap="Queries Per Second and Time to Build Indexes for 20-dimension Euclidean", fig.width=6, fig.height=3}
library(knitr)
times <- data.frame(Algorithm=c("BallTree(leaf_size=10)", "BallTree(leaf_size=100)", "BallTree(leaf_size=1000)",
                            "TrilaterationExpand2", "BruteForce", "Trilateration", "TrilaterationExpand"),
                    QPS=c(3974.894, 2561.925, 2224.275, 1411.711, 670.163, 25.971, 401.899))
times <- times[order(times$Algorithm),]
# kable(times, caption="Queries Per Second and Time to Build Indexes for 20-dimension Euclidean")

p <- ggplot(data=times, aes(x=Algorithm, y=QPS)) +
  geom_bar(stat="identity", col='black', fill='grey') +
  coord_flip()
p
```

### Glove 25-dimension Angular distance

The GloVe ("Global Vectors for Word Representation") dataset "is an unsupervised learning algorithm for obtaining vector representations for words. Training is performed on aggregated global word-word co-occurrence statistics from a corpus, and the resulting representations showcase interesting linear substructures of the word vector space" according to the authors. [@pennington2014glove]

It is one of the built-in datasets in the ann-benchmark tool.  Under the hood, this isusing a euclidean distance function, once the angular coordinates are transformed, so the relative performance is similar to the euclidean dataset.

Of note, we include only results that were full (not approximate) nearest neighbor solutions.

These results include the addition of the FAISS algorithm, one of the graph index based NN solvers which came out of Facebook Research in recent years.[@FAISS2017]  On these datasets, FAISS is a beast, but unfortunately it is a very highly tuned GPU-based implementation which makes it difficult to adapt to unsupported distance functions, such as the Geodesic we target with Trilateration.  We leave it here for information, but are unable to compare it on our core task.

Note that our performance has suffered again similarly to with the random euclidean dataset.  Trilateration is extremely slow; as is TrilaterationExpand.  TrilaterationExpand2 beats BruteForce and is somewhat shy of the Tree-based algorithms.  But we are not competitive in this space.

```{html echo=FALSE}
<!--
writing output to results/glove-25-angular.png
  2:                             FaissIVF(n_list=64, n_probe=200)        1.000       35.085
  4:                           FaissIVF(n_list=1024, n_probe=200)        1.000      164.781
  8:                             FaissIVF(n_list=32, n_probe=200)        1.000       35.108
 10:                            FaissIVF(n_list=128, n_probe=100)        1.000       43.439
 16:                             FaissIVF(n_list=64, n_probe=100)        1.000       35.081
 18:                            FaissIVF(n_list=256, n_probe=200)        1.000       44.869
 25:                             FaissIVF(n_list=128, n_probe=50)        1.000       83.435
 31:                            FaissIVF(n_list=512, n_probe=200)        1.000       87.226
 39:                            FaissIVF(n_list=256, n_probe=100)        1.000       88.562
 41:                             FaissIVF(n_list=32, n_probe=100)        1.000       35.103
 47:                            FaissIVF(n_list=128, n_probe=200)        1.000       34.741
 48:                              FaissIVF(n_list=64, n_probe=50)        1.000       44.832
 52:                              FaissIVF(n_list=32, n_probe=50)        1.000       35.020
 55:                                      BallTree(leaf_size=200)        1.000        8.084
 56:                                                 BruteForce()        1.000        3.765
 99:                                                Trilateration        1.000        0.010  # Estimate from 1000 queries
 99:                                          TrilaterationExpand        1.000        1.090
 54:                                         TrilaterationExpand2        1.000        6.125
 -->
```



```{r glove-25-angular, echo=FALSE, fig.cap="Queries Per Second and Time to Build Indexes for glove-25-angular", fig.width=6, fig.height=3}
times <- data.frame(Algorithm=c(
"FaissIVF(n_list=64, n_probe=200)",
"FaissIVF(n_list=1024, n_probe=200)",
"FaissIVF(n_list=32, n_probe=200)",
"FaissIVF(n_list=128, n_probe=100)",
"FaissIVF(n_list=64, n_probe=100)",
"FaissIVF(n_list=256, n_probe=200)",
"FaissIVF(n_list=128, n_probe=50)",
"FaissIVF(n_list=512, n_probe=200)",
"FaissIVF(n_list=256, n_probe=100)",
"FaissIVF(n_list=32, n_probe=100)",
"FaissIVF(n_list=128, n_probe=200)",
"FaissIVF(n_list=64, n_probe=50)",
"FaissIVF(n_list=32, n_probe=50)",
"TrilaterationExpand2",
"BallTree(leaf_size=200)",
"BruteForce()",
"Trilateration",
"TrilaterationExpand"),
QPS = c(35.085, 164.781, 35.108, 43.439, 35.081, 44.869,
83.435, 87.226, 88.562, 35.103, 34.741, 44.832, 35.020,
6.125, 8.084, 3.765, 0.010, 1.000))
# kable(times, caption="Queries Per Second and Time to Build Indexes for glove-25-angular")
p <- ggplot(data=times, aes(x=Algorithm, y=QPS)) +
  geom_bar(stat="identity", col='black', fill='grey') +
  # geom_text(position=position_dodge(width = 0.5)) +
  coord_flip()
p
```