# Introducing the Multilateration Index

## Multilateration Index – General Definition 

Given an n-dimensional metric space $(M, d)$ comprising universe of points $M$ in the space and a distance function $d(x, y)$ which respects the triangle inequality, a typical point $X$ in the coordinate system will be described by coordinates ${x_1, x_2, …, x_n}$, which, typically, represents the decomposition of a vector $V$ from an "origin" point $O: {0, 0, … ,0}$ to $X$ into orthogonal vectors $\{x1,0,..,0\}, \{0,x2,0,..,0\}, …, \{0,0,..,0,xn\}$ along each of the n dimensional axes of the space.

The Multilateration of points in the space requires $n+1$ fixed reference points $F_p$ ($p$ from $1$ to $n+1$), which contains no subset of any length $m$ which all lie on the same $m-2$ hyperplane.  (i.e. in a 3d coordinate system, with four reference points, the four points cannot lie on the same plane, no three points can lie on the same line, and, trivially, no two points can be the same).  The Multilateration Coordinate $X'$ for the point $X$ is then: $X'=\{t_1, t_2, …, t_{n+1}\}$ where $t_i$ is the distance $d(X, F_i)$ (in units applicable to the system).

A "Multilateration Index" is a data structure which, for a given set of points $p \in M$ stores the Multilateration Coordinates.  Depending on implementation and use, these can be stored in a single structure sorted by one of $t_i$, or in multiple structures sorted individually by each $t_i$, or some other variant.  See [Multilateration NN Algorithms] and [Multilateration NA Algorithms] for our specific implementations.

## 2-D Bounded Example 

Consider a 2-dimensional grid – a flattened map, a video game map, or any mathematical $x-y$ coordinate grid with boundaries.  WOLOG in this example consider the two-dimensional Euclidean space $M=\mathbb{R^2}$ and bounded by $x,y\ \epsilon\  \{0..100\}$.  Also, let us use the standard Euclidean distance function for $d$.  This is, trivially, a valid metric space.

Since the space has dimension $n=2$, we need $3$ fixed points $F_p$.  While the Geospatial example on Earth has a specific prescription for the fixed points, an arbitrary space does not.  We therefore prescribe the following construction for bounded spaces: 

Construct a circle (hypersphere for other dimensions) with the largest area inscribable in the space.  In this example, that will be the circle centered at $(50,50)$ with radius $r=50$. 

Select the point at which the circle touches the boundary at the first dimension (for spaces with uneven boundary ratios, select the point at which the circle touches the earliest boundary $x_i$).  Such a point is guaranteed to exist since the circle is largest (if it does not, then the circle can be expanded since there is space between every point on the circle and an axis, and it is not a largest possible circle).

From this point, create a regular $n+1$-gon (triangle here) which touches the circle at $n+1$ points.  These are the points we will use as $F_p$.  They are, by construction, not all co-linear (or in general do not all exist on the same $n$-dimensional hyperplane) satisfying our requirement [proof]. 

The point $y=0$, $x=50$ is the first point of the equilateral triangle.  The slope of the triangle's line is $tan(\frac{pi}{3})$, so setting the equation of the circle: 

$(x-50)^2+(y-50)^2=50^2$ equal to the lines: $y=tan(\frac{\pi}{3})(x-50)$ gives $x=25(2+\sqrt{3})$ on the right  and $y=tan(\frac{-\pi}{3})(x-50)$ gives $x=-25(\sqrt{3}-2)$ on the left, and of course the original $(0,50)$ point.  Applying $x$ to our earlier equations for $y$ we get a final set of three points: 

$$F_1 = (x=50,y=0)$$

$$F_2 = (x=25(2+\sqrt{3}), y=tan(\frac{\pi}{3})((25(2+\sqrt{3}))-50)$$

$$F_3 = (x=-25(\sqrt{3}-2), y=tan(\frac{-\pi}{3})((-25(\sqrt{3}-2))-50) $$

```{r trianglePlot, fig.align="center", echo=FALSE}
library(plotrix)

plot(x=NULL, y=NULL, xlim=c(0,100), ylim=c(0,100), asp = 1,
     xlab='X', ylab='Y',
     main = "Example calculation of reference points in 2d area")
draw.circle(x = 50, y = 50, radius = 50)
abline(h=0)
abline(h=100)
abline(v=0)
abline(v=100)

# Draw point
draw.circle(x = 50, y = 0, radius = 1, lwd = 1, col="red")

# Triangle points:
myTri <- data.frame(id=1, x=50, y=0)
myTri <- rbind(myTri, c(2, 25*(2+sqrt(3)), tan(pi/3)*((25*(2+sqrt(3)))-50)))
myTri <- rbind(myTri, c(3, -25*(sqrt(3)-2), tan(-pi/3)*((-25*(sqrt(3)-2))-50)))
polygon(myTri[,2], myTri[,3], border="blue")

# abline(a=tan(pi/3)*(-50), tan(pi/3))


# Draw point
draw.circle(x = 50, y = 0, radius = 1, lwd = 1, col="red")
draw.circle(x = 25*(2+sqrt(3)), y = tan(pi/3)*((25*(2+sqrt(3)))-50), radius = 1, lwd = 1, col="red")
draw.circle(x = -25*(sqrt(3)-2), y = tan(-pi/3)*((-25*(sqrt(3)-2))-50), radius = 1, lwd = 1, col="red")
```

Remember, any three non-colinear points will do, but this construction spaces them fairly evenly throughout the space, which may be beneficial later* [Add section (reference) with discussions of precision and examples where reference points are very near one another]. 

The trilateration of any given point X in the space, now, is given by: 

$$T(X) = {d(F_1, X), d(F_2, X), d(F_3,X)} $$

That is, the set of (three) distances $d$ from $X$ to $F_1$, $F_2$, and $F_3$ respectively. 

#### 10 Random Points

As a quick example of the trilateration calculations, we use a basic collection of 10 data points:

```{r rand10points, echo=FALSE}
  library(knitr)
  myDist <- function(x, y) {
    # x and y must be 2D points -- (x,y) tuples i.e.: x <- c(1,2) y <- c(3,4)
    d <- sqrt((x[[1]]-y[[1]])^2+(x[[2]]-y[[2]])^2)
    return(d)
  }

  # Function to generate lots of random points
  makepoints <- function(n, min, max) {
    set.seed(1729)
    myPoints <- data.frame(x = runif(n = n, min = min, max = max),
                           y = runif(n = n, min = min, max = max))
    return(myPoints)
  }

  # 10 Random Points (seed ensures reproducibility):
  myPoints <- makepoints(10, 0, 100)
  kable(myPoints, caption = "10 Random Points")
```


The trilateration of those points, that is, the three points $d_1, d_2, d_3 = {d(F_1, X), d(F_2, X), d(F_3,X)}$ are (next to the respective $x_n$):

```{r distanceexample, echo=FALSE}
  myTrilat <- data.frame(d1=NULL, d2=NULL, d3=NULL)
  for(i in 1:nrow(myPoints)) {
    myTrilat[i, 'd1'] <- myDist(c(myPoints[i,'x'], myPoints[i, 'y']), c(myTri[1,'x'], myTri[1,'y']))
    myTrilat[i, 'd2'] <- myDist(c(myPoints[i,'x'], myPoints[i, 'y']), c(myTri[2,'x'], myTri[2,'y']))
    myTrilat[i, 'd3'] <- myDist(c(myPoints[i,'x'], myPoints[i, 'y']), c(myTri[3,'x'], myTri[3,'y']))
  }

  myTrilateration <- cbind(myPoints, myTrilat)
  rm(myTrilat)  #twas temporary
  kable(myTrilateration, caption="Multilateration Index of 10 Random Points WRT Fixed points F")
```

Note that we do not need to continue to store the original latitude and longitude.  We can convert the three $d_n$ distances back to Latitude and Longitude within some $\epsilon$ based on the available precision.  Geospatial coordinates in Latitude and Longitude with six digits of precision are accurate to within $<1\ meter$, and 8 digits is accuract to within $<1\ centimeter$, although this varies based on the latitude and longitude itself; latitudes closer to the equator are less accurate than those at the poles.  The distance values $d_x$ are more predictable, since they measure distances directly.  While the units in this sample are arbitrary, $F(x)$ in a real geospatial example could be in kilometers, so three decimal digits would precisely relate to $1\ meter$, and so on.  This is one reason that we will later examine using the trilateration values as an outright replacement for Longitude and Latitide, and this feature is important when considering storage requirements for this data in large real-world database applications.

For now, continuing with the example, those $10$ points are shown here in blue with the three reference points $F_1, F_2, F_3$ in red:

```{r plot10Points, fig.align="center", echo=FALSE}

  plot(x=NULL, y=NULL, xlim=c(0,100), ylim=c(0,100), asp = 1,
       xlab='X', ylab='Y', main = "Sample Reference and Data Points")
  points(myTrilateration, col="blue", pch=16, asp=1)
  text(myTrilateration, pos = 1)
  
  draw.circle(x = myTri[1,'x'], y = myTri[1,'y'], radius = 1, lwd = 1, col="red")
  draw.circle(x = myTri[2,'x'], y = myTri[2,'y'], radius = 1, lwd = 1, col="red")
  draw.circle(x = myTri[3,'x'], y = myTri[3,'y'], radius = 1, lwd = 1, col="red")
  text(myTri[1, 'x'], myTri[1, 'y'], labels = 'F1', pos=2)
  text(myTri[2, 'x'], myTri[2, 'y'], labels = 'F2', pos=1)
  text(myTri[3, 'x'], myTri[3, 'y'], labels = 'F3', pos=2)
```

To help understand the above values, the following chart shows the distances for points $x_2$ and $x_10$ above.  Specifically, the distances $d_1$ from point $F_1$ are shown as arcs in red, the distances $d_2$ from point $F_2$ in blue, and $d_3$ from point $F_3$ in green.  

```{r distanceExamples, fig.align="center", echo=FALSE}
# Arcs around two points:
  
  plot(x=NULL, y=NULL, xlim=c(0,100), ylim=c(0,100), asp = 1,
       xlab='X', ylab='Y', main = "Distance Arcs to Two Sample Points")
  points(myTrilateration[c(2,10),'x'], myTrilateration[c(2,10),'y'], col="blue", pch=16, asp=1)
  text(myTrilateration[2,'x'], myTrilateration[2,'y'], labels = 'x2', pos=4)
  text(myTrilateration[10,'x'], myTrilateration[10,'y'], labels = 'x10', pos=2)
  
  draw.circle(x = myTri[1,'x'], y = myTri[1,'y'], radius = 1, lwd = 1, col="red")
  draw.circle(x = myTri[2,'x'], y = myTri[2,'y'], radius = 1, lwd = 1, col="red")
  draw.circle(x = myTri[3,'x'], y = myTri[3,'y'], radius = 1, lwd = 1, col="red")
  text(myTri[1, 'x'], myTri[1, 'y'], labels = 'F1', pos=2)
  text(myTri[2, 'x'], myTri[2, 'y'], labels = 'F2', pos=1)
  text(myTri[3, 'x'], myTri[3, 'y'], labels = 'F3', pos=2)
  
  
  draw.circle(x=myTri[1,'x'], y=myTri[1,'y'], radius=myTrilateration[2,'d1'], border="red")
  draw.circle(x=myTri[1,'x'], y=myTri[1,'y'], radius=myTrilateration[10,'d1'], border = "red")

  draw.circle(x=myTri[2,'x'], y=myTri[2,'y'], radius=myTrilateration[2,'d2'], border="blue")
  draw.circle(x=myTri[2,'x'], y=myTri[2,'y'], radius=myTrilateration[10,'d2'], border="blue")

  draw.circle(x=myTri[3,'x'], y=myTri[3,'y'], radius=myTrilateration[2,'d3'], border="darkgreen")
  draw.circle(x=myTri[3,'x'], y=myTri[3,'y'], radius=myTrilateration[10,'d3'], border="darkgreen")
```


### Geospatial Trilateration Index

Applying this to real sample points; we propose a standard geospatial trilateration index to allow for comparability across data sets by setting the following specific initial reference points on the globe: 

Point 1: $90.000000, 0.000000$ (The geographic north pole) 

Point 2: $38.260000, -85.760000$ (Louisville, KY on the Ohio River) 

Point 3: $-19.22000, 159.93000$ (Sandy Island, New Caledonia)

Optional Point 4: $-9.42000, 46.33000$ (Aldabra)

Optional Point 5: $-48.87000, -123.39000$ (Point Nemo)

Note that the reference points are defined precisely, as exact latitude and longitude to stated decimals (all remaining decimal points are 0).  This is to avoid confusion, and why the derivation of the points is immaterial (Point Nemo, for example is actually at a nearby location requiring more than two digits of precision).

Only three points are required for trilateration (literally; thus the "tri" prefix of the term), but we include 5 points to explore the pros and cons of n-fold geodesic indexing for higher values of n.  These optional points turn out to serve no real purpose, as we were unable to improve performance with the use of additional points, however we leave them for posterity and for consistency in future experimentation.  We note that Sandy Island, New Caledonia is not a real place [@Seton2013], so there may be value in replacing it in some situations.  Point Nemo was selected as being the furthest point from any piece of land. [@Barnes2020]  The thought being that if distances were more commonly measured on land, perhaps an anchor point like this may provide more uniformity in distances.  In any case these alternatives proved to be of no particular value.
