---
title: "MULTILATERATION INDEX"
author: |
  | Chip Lynch
  |
  | A Thesis Submitted to the University of Louisville
  | J.B. Speed School of Engineering
  | for the Degree of Master of Science in Computer Science
date: "07/20/2021"
output:
  pdf_document:
    number_sections: true
    extra_dependencies: ["float"]
  html_document: default
institute: University of Louisville
bibliography: references.bib
geometry: "left=0cm,right=0cm,top=1in,bottom=0cm"
indent: false
toc-title: "TABLE OF CONTENTS"
header-includes: 
  - \renewcommand\refname{References}
  - \usepackage{placeins}
  - \usepackage{setspace}
  - \usepackage{ragged2e}
  - \AtBeginDocument{\pagenumbering{roman}}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'H')
# knitr::opts_chunk$set(fig.pos = 'H')
# knitr::opts_chunk$set(fig.pos = "!H", out.extra = "")
# Keep figures from floating - thank you https://gist.github.com/burchill/8873d2ade156b27e92a238a774ce2758:
knitr::knit_hooks$set(plot = function (x, options) {
  float_correct <- function(f, y, opts)  {
    if (is.null(opts$regfloat) || opts$regfloat==FALSE)
      paste0(f(y, opts), "\n\n\\FloatBarrier\n")
    else
      f(y, opts)
  }
  if (!is.null(options$out.width) || !is.null(options$out.height) ||
      !is.null(options$out.extra) || options$fig.align != "default" ||
      !is.null(options$fig.subcap)) {
    if (is.null(options$fig.scap))
      options$fig.scap = NA
    return(float_correct(knitr:::hook_plot_tex, x, options))
  }
  return(float_correct(knitr:::hook_plot_md_base, x, options))
})
```


```{r include=TRUE, fig.align="center", echo=FALSE, out.width="100%", out.height="100%"}
knitr::include_graphics("./Signature_Sheet_Chip Lynch.pdf")
```

\newgeometry{left=1.5in,right=1in,top=1in,bottom=1in}

```{r child = 'Acknowledgements.rmd'}
```

\newpage
\doublespacing
```{r child = 'Abstract.rmd'}
```

\singlespacing

\newpage
$$\\[0.5in]$$
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\renewcommand\contentsname{TABLE OF CONTENTS}
\tableofcontents

\newpage

\renewcommand\listtablename{LIST OF TABLES}
\listoftables

\renewcommand\listfigurename{LIST OF FIGURES}
\listoffigures

\newpage

\doublespacing

\cleardoublepage
\pagenumbering{arabic}

\newpage
```{r child = 'Introduction.rmd'}
```

\newpage

```{r child = 'ReviewOfLiterature.rmd'}
```

\newpage

```{r child = 'NetworkAdequacyDefinition.rmd'}
```

\newpage

```{r child = 'MultilaterationIndex.rmd'}
```


```{r child = 'MathTheory.rmd'}
```

\newpage

```{r child = 'Experiment.rmd'}
```

\newpage

```{r child = 'ConclusionsAndFutureWork.rmd'}
```


\newpage

```{r child = 'Appendix.rmd'}
```
